---
title: "Hotel"
author: "Nik Agarwal"
date: "4/14/2020"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
library(tidyverse)

# load dataset
hotels <- vroom::vroom(
  file = "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv",
  na = c("", " ", "NA", "NULL"))
```


```{r df_preview}
glimpse(hotels)
```


```{r null_chk}
hotels %>%
  purrr::map_df(function(x) sum(is.na(x))) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = 'missing'
  ) %>%
  dplyr::arrange(desc(missing)) %>%
  mutate(
    percent_missing = round(missing / nrow(hotels),2)
  ) %>%
  filter(missing > 0)
```

Conclusions
- We can remove the variable <company> from the entire training.
- We can consider removing the variable <agent> since 14% of the observations are missing
- We can consider simply remove the missing values for country

Actions
- Remove the variable company
- Replace missing values in variable agent with 9999 (indicating that it was an unknown agent)
- drop all rows that are missing country and missing children (losing 492 rows)

**check to see response variable**

```{r hotel_type}

hotels %>%
  group_by(hotel) %>%
  count() %>%
  rename(obs_count = n) %>%
  mutate(
    pct = round(obs_count / nrow(hotels),2)
  )

```

```{r response_var}

hotels %>%
  group_by(is_canceled) %>%
  count() %>%
  rename(obs_count = n) %>%
  mutate(
    pct = round(obs_count / nrow(hotels),2)
  )

```
```{r}
hotels %>%
  group_by(hotel, is_canceled) %>%
  count() %>%
  rename(obs_count = n) %>%
  pivot_wider(
    names_from = is_canceled,
    values_from = obs_count,
    values_fill = list(obs_count = 0)
  ) %>%
  rename(not_cancelled = `0`, cancelled = `1`)
```
```{r}
hotels %>%
  group_by(hotel, is_canceled) %>%
  count() %>%
  rename(obs_count = n) %>%
  mutate(
    pct = round(obs_count / nrow(hotels),2)
  ) %>%
  select(-obs_count) %>%
  pivot_wider(
    names_from = is_canceled,
    values_from = pct,
    values_fill = list(obs_count = 0)
  ) %>%
  rename(not_cancelled = `0`, cancelled = `1`)
```

```{r}

total_city <- hotels %>% filter(hotel == 'City Hotel') %>% count() %>% pull()
total_resort <- hotels %>% filter(hotel == 'Resort Hotel') %>% count() %>% pull()

hotels %>%
  group_by(hotel, is_canceled) %>%
  count() %>%
  rename(obs_count = n) %>%
  mutate(
    total_obs = case_when(
      hotel == 'City Hotel' ~ total_city,
      TRUE ~ total_resort
    ),
    pct = round(obs_count / total_obs,2)
  ) %>%
  pivot_wider(
    id_cols = hotel,
    names_from = is_canceled,
    values_from = pct
  ) %>%
  rename(cancelled = `1`, not_cancelled = `0`)

```
We will still make 1 model even though there are two types of hotels.

## Feature Engineering

- Replace missing values in travel agent column with 9999
- discard rows that have missing values for country & children
- create feature that looks at the day of week of check-in (M, T, W, etc.)
- create feature that tells the duration of stay (planned or actual)
- create feature that tells us total guests
- total cost of stay
- create feature if meal was included in reservation or not

```{r create_newdf}
chorizo <- hotels
```



```{r replace_agent}


chorizo <- chorizo %>%
  mutate(
    agent = ifelse(is.na(agent), 9999, agent)
  )

```


```{r remove_company}

chorizo <- chorizo %>%
  select(-company)

```


```{r remove_rows}

chorizo <- chorizo %>%
  drop_na()

```


```{r confirm_nomissingvals}

chorizo %>%
  map_df(function(x) sum(is.na(x))) %>%
  pivot_longer(
    cols = everything(),
    names_to = 'variable',
    values_to = "missing"
  ) %>%
  arrange(desc(missing))

```


```{r dayofweek}
chorizo <- chorizo %>%
  mutate(
    reservation_date = lubridate::ymd(paste0(arrival_date_year,"-",arrival_date_month,"-",arrival_date_day_of_month)),
    day_of_week = lubridate::wday(reservation_date, label = T)
  )
```


```{r stay_duration}

chorizo <- chorizo %>%
 mutate(
    stay_duration = stays_in_week_nights + stays_in_weekend_nights
  )

```


```{r total_guests}

chorizo <- chorizo %>%
  mutate(
    total_guests = adults + children + babies
  )

```


```{r total_cost_of_stay}
chorizo <- chorizo %>%
  mutate(
    total_cost = stay_duration * adr
  ) %>%
  filter(total_cost > 0)
```


```{r meal_included}

chorizo <- chorizo %>%
  mutate(
    meal_included = case_when(meal %in% c('SC','Undedfined') ~ 0, TRUE ~ 1)
  )

```


Let's get ride of extra columns now

```{r delete_extra_columns}

chorizo <- chorizo %>%
  select(-arrival_date_year, -arrival_date_month, -arrival_date_week_number, -arrival_date_day_of_month, -stays_in_weekend_nights, -stays_in_week_nights, -adults, -children, -babies, -reservation_status, -reservation_status_date, -adr, -meal, -distribution_channel)

```


# create training, validation, and test data sets


```{r splits}
set.seed(8837)
splits <- rsample::initial_split(chorizo, prop = 0.8, strata = is_canceled)

df_train <- splits %>% rsample::training()
df_test <- splits %>% rsample::testing()

# split the training dataset further

set.seed(3362)
vsplit <- rsample::initial_split(df_train, prop = 0.8, strata = is_canceled)
df_train <- vsplit %>% rsample::training()
df_valid <- vsplit %>% rsample::testing()


```

